\vspace{2cm} \LARGE \bf Seasonal influenza circulation patterns and projections for 2016-2017

### 2016-09-21 auspice/reports/tex/sep-2016.tex

\date{September 20, 2016}

@@

We base our primary analysis on a set of viruses collected between Sep 2014 and Aug 2016, comprising approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible (Fig.\ \ref{H3N2_counts}). This equilibration attempts to collect equal samples from Africa, China, Europe, Japan/South Korea, North America, Oceania, South America, South Asia, Southeast Asia and West Asia. In the following analysis we collapse samples from China, South Asia, Southeast Asia, Japan and Korea into a single region referred to here as ``Asia'', resulting in Asia possessing greater sample counts than North America or Europe. The only month that significantly departs from equitable sampling is Aug 2016 with 38 viruses, primarily from Europe and North America. We subsample to 100 viruses per month and not more to keep sample counts as equitable as possible across space and time. Repeating mutation and clade frequency calculations with up to 200 viruses per month and region yields similar results (see below).

@@

The 142K/197R clade first emerged around June 2015 and rose to nearly 20\% global frequency in Nov 2015 (Fig.\ \ref{H3N2_mutations}). However, it's declined throughout 2016. We doubt this is a competitive virus. The 197K clade emerged in late 2015 and has slowly grown in frequency since. We estimate that it now comprises $\sim$14\% of H3N2 viruses. However, 197K is at extremely low frequency in Asia with an estimated present-day frequency $\sim$2\%. On the other hand, clade 171K viruses have done remarkably well throughout 2016 and now comprise an estimated 69\% of currently circulating H3N2 viruses. This steady and rapid increase is strongly suggestive of an adaptive origin. Notably, 171K emerged and spread first in Asia, reaching nearly 80\% frequency in April 2016. Higher frequency of 171K in Asia is expected to spread to the rest of the world given historic geographic observations \cite{bedford2015global}. Within 171K, the variant 121K has spread. However, it doesn't appear to be spreading more rapidly than its parent 171K clade. Phylogenetic patterns suggest 171K as driver rather than 121K. Still, the rate of increase of 121K suggests that a sizable fraction of 2016-2017 viruses will be comprised of 121K.

@@

The clade and mutation frequencies discussed above were based on a limited sequence subsample with an equitable geographical distribution. More accurate region specific trajectories can be estimated using all sequence data available in GISAID. We repeated the clade and mutation frequency estimation using up to 200 sequences per month and region (Fig.\ \ref{H3N2_counts_all}).

>>>>

\date{September 21, 2016}

@@

We base our primary analysis on a set of viruses collected between Sep 2014 and Aug 2016, comprising approximately 100 viruses per month where available and seeking to equilibrate sample counts geographically where possible (Fig.\ \ref{H3N2_counts}). This equilibration attempts to collect equal samples from Africa, China, Europe, Japan/South Korea, North America, Oceania, South America, South Asia, Southeast Asia and West Asia. In the following analysis we collapse samples from China, South Asia, Southeast Asia, Japan and Korea into a single region referred to here as ``Asia'', resulting in Asia possessing greater sample counts than North America or Europe. The only month that significantly departs from equitable sampling is Aug 2016 with 38 viruses, primarily from Europe and North America. We subsample to 100 viruses per month and not more to keep sample counts as equitable as possible across space and time. Repeating mutation and clade frequency calculations with up to 1200 viruses per month yields similar results (see below).

@@

The 142K/197R clade first emerged around June 2015 and rose to nearly 20\% global frequency in Nov 2015 (Fig.\ \ref{H3N2_mutations}). However, it's declined throughout 2016. We doubt this is a competitive virus based on current clade success, but site 142 mutated several times within 3c2a to 142G or 142K suggesting that mutating this position might enable the virus to escape preexisting immunity at a fitness cost; it warrants continued observation. The 197K clade emerged in late 2015 and has slowly grown in frequency since. We estimate that it now comprises $\sim$14\% of H3N2 viruses. However, 197K is at extremely low frequency in Asia with an estimated present-day frequency $\sim$2\%. On the other hand, clade 171K viruses have done remarkably well throughout 2016 and now comprise an estimated 69\% of currently circulating H3N2 viruses. This steady and rapid increase is strongly suggestive of an adaptive origin. Notably, 171K emerged and spread first in Asia, reaching nearly 80\% frequency in April 2016. Higher frequency of 171K in Asia is expected to spread to the rest of the world given historic geographic observations \cite{bedford2015global}. Within 171K, the variant 121K has spread. However, it doesn't appear to be spreading more rapidly than its parent 171K clade. Phylogenetic patterns suggest 171K as driver rather than 121K. Still, the rate of increase of 121K suggests that a sizable fraction of 2016-2017 viruses will be comprised of 121K.

@@

The clade and mutation frequencies discussed above were based on a limited sequence subsample with an equitable geographical distribution. More accurate region specific trajectories can be estimated using all sequence data available in GISAID. We repeated the clade and mutation frequency estimation using up to 500 sequences per month and region (Fig.\ \ref{H3N2_counts_all}).
